% \documentclass[openacc]{rsproca_new}
\documentclass[a4paper]{article}
% \titlehead{Research}

\usepackage[margin=1in]{geometry}
\usepackage{hyperref}
\usepackage{color}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{amssymb}
\numberwithin{equation}{section}
\usepackage[backend=bibtex, style=numeric-comp, sorting=none]{biblatex}
\DeclareFieldFormat{labelnumberwidth}{[#1] }
\addbibresource{main.bib}

\begin{document}

\title{Statistical mechanics of Hamiltonian curl-force systems}

\author{%%%% Author details
Omer Chor$^{1}$, Alja\v{z} Godec$^{2}$ and Oren Raz$^{1}$}

%%%%%%%%% Insert author address here
\date{
$^{1}$Department of Physics of Complex Systems, Weizmann Institute of Science, Rehovot, 76100, Israel
\\
$^{2}$ \textcolor{red}{Add Alja\v{z}'s affiliation}\\[2ex]
\today}

\maketitle

%%%% Subject entries to be placed here %%%%
% \subject{xxxxx, xxxxx, xxxx}

%%%% Keyword entries to be placed here %%%%
% \keywords{xxxx, xxxx, xxxx}

%%%% Insert corresponding author and its email address}
% \corres{Insert corresponding author name\\
% \email{xxx@xxxx.xx.xx}}

% \date{\today}

\begin{abstract}
We investigate the statistical mechanics of a particle subject to curl-forces---reversible, velocity independent non-conservative forces with non-vanishing curl---that can be described by a Hamiltonian with an anisotropic kinetic energy term. We show that generally they do not relax to the corresponding Boltzmann distribution when coupled to a heat bath, and therefore the Hamiltonian structure cannot be used to easily find their nonequilibrium steady state. By applying a canonical transformation, the system can be mapped to a system without curl-forces that is instead coupled to multiple heat baths at different temperatures. For quadratic potentials, this mapping recovers the underdamped Brownian gyrator, providing a new perspective on the origin of its rotational currents as the result of deterministic, non-conservative forces. The same analysis is also applied to another variation of the Brownian gyrator, demonstrating its generality.
\end{abstract}

\maketitle

\section{Introduction}
\label{sec:Intro}
\input{Sec_Introduction}

\section{Hamiltonian curl-forces}

\label{sec:HCF}
\input{Sec_CurlForce}

\section{Statistical mechanics}
\label{sec:StochDyn}
\input{Sec_StochasticDynamics}

\section{Applications}
\label{Sec:Applications}
\input{Sec_Applications}
\input{Sec_discussion}

 \paragraph{Acknowledgments}
\textcolor{red}{Add Alja\v{z}'s funding and Omer's (if any should be listed)}
OR acknowledges financial support from the ISF, grant no. 232/23,  by the Minerva foundation with funding from the Federal German Ministry for Education and Research and by the Lower Saxony - Israel collaboration grant. We would like to thank Roi Holtzman and Shahaf Aharony Shapira for helpful discussions.
 
% \bibliographystyle{RS}
% \bibliography{main}
\printbibliography

\end{document}